 

### 2021-01-16 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory. This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and Co2 which influences leaf traits, thermoregulation and metabolism along the gradient, and with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].

The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019]. Conversly, available water facilitates leaf transpiration through open stomata, however increased relative humdity in tropical regions indirectly caused by increased temperature, reduces transpiration leading to high leaf temperatures in the canopy [Tibbitts 1979; Perez and Feeley 2018].

Frequent droughts related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020]. Larger trees suffer hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al. 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a]. Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020]. However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates [REF].

@@

While there is evidence for temperature driven shifts in forest dynamics, leaf-level metabolic and performance changes influenced by biophysical conditions and warming in the overstory and understory is still unclear. Thus, it is becoming important to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy. This will provide a pathway for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions with additional feedbacks to forest structure and microclimates.

@@

Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [Rosendaal et al. 2006]. After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [Keenan and Niinemets et al.2016; @niinemetsLeafAgeDependent2016a]. A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR, and the ratio of red to far red radiation [R:FR] intercepting the leaf surface. This enables acclimation or avoidance responses through the development of necessary leaf traits [Casal, 2013]. As a result, degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], which ultimately also influences leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al. 2016].

@@

Additionally, shade leaves in the understory can be broken into three categories; 1. shade leaves of canopy trees--top of the tree is sun exposed, bottom is shaded. 2. seedlings/saplings of canopy trees that will eventually, if they survive, reach the canopy. 3 understory plants that complete their lifecycle in the shaded understory. These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered [refs, -Martijn Slot]

@@

Taller trees in dense forests tend to be exposed to greater irradiance but maintain shaded leaves in canopy interior and lower canopy compared to shorter understory trees that are shaded by overstory. As a result, the sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, hairier, with thicker cuticle and greater lobation than shade leaves
[Rozendaal et al. 2006, Mathur et al. 2018, more REFS].

However, the combination of stressors experienced by taller trees' foliage are different from that experienced by understory trees. For example, overstory sun leaves in particular experience greater loads of PAR, $VPD$, elevated temperatures, higher wind speed, etc.. whereas for understory leaves are subjected to --shade or limitation of light, niche competition and herbivory. Overstory sun leaves are subjected to overheating pressures, thus through developed traits of increased stomatal density, transpiration and water supply along with increased convective heatloss by reducing boundary layer resistance, sun leaves are able to maintain optimum temperature for photosynthesis and metabolic processes. Additionally, overstory sun leaves tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Adaptive traits in understory enable light capture efficiency while maintaing larger width leaves with comparitevely fewer stomtal density as a result of available water, lesser $VPD$, and wind speed. "Spending their lifecycle in the shade means they have different pressures than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well."[-Martijn]. Those differences in traits are likely to affect their thermal properties as well. Leaves act as windows to plant performance, enhancing plant adaptation to changing environment, thus leaf traits acclimated to prior stressors have been suggested to enable whole plants to respond differently to forthcoming stress [Rozendaal et al. 2006; @niinemetsResponsesForestTrees2010a]

@@

With increasing height and age, competition for light decreases as availability  of light increases in dense forests. However, depending on growth strategies of species occupying the understory, leaf traits might differ due to species inherent light requirements. Early successional species are considered more shade intolerant than late successional species that are more shade tolerant. Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities in shade tolerators than in intolerators [Niinemets, 2006]. Shade intolerant species experience greater heterogeniety in light during ontogeny [why?-nidhi],thus LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light [H. Poorter, U. Niinemets, L. Poorter, I. J. Wright, R. Villar, Causes and consequences of variation in leaf mass per area (LMA): A meta-analysis. New Phytol. 182, 565–588 (2009).].Whereas, Rozendaal et al. 2006 found no significant trait differences with ontogeny and adult stature trees. Koike et al. 2001 has more info about this.

@@

Leaf temperature is generally associated to increase with leaf area ($LA$) [ Wright et al. 2019]. Thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017]. With increase in efffective leaf width, boundary layer thickness and resistance increases along with stomtal resistance [@leighInfluenceLeafSize2017; Wright et al. 2019; @fauset_differences_2018].This lowers evaporation from stomatal pores, and in turn traspirational cooling in large leaves [@fauset_differences_2018]. Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017]. Using leaf energy balance to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al. 2016 found leaf width to be a dominant factor in deterimining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$. Leaf width and characteristic dimension is generally greater in shaded leaves along the vertical gradient than exposed upper canopy [ref]. Additionally, leaf characteristics such as lobes, leaf elongation and serration break the entirity of leaf dimension by enhancing a smaller characteristic dimension, all facilitate in dissapating heat more efficiently. Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [Vogel 1970; Roth-Nebelsick 2001;Schuepp 1993; Sack et al. 2006].However, Perez and Feeley et al. 2020 showed only a slight negetive effect on leaf temperature with leaf width.

@@

Thermal time constant, a leaf trait, quantifies heat flux in a leaf through incorporating heat stored versus heat exchanged with the environment, with relation to time. It comprises of several underlying traits -- leaf area, LMA, leaf width, leaf dry matter content and stomatal conductance [@michaletzEnergeticCarbonEconomic2016]. Typically, larger leaves with greater leaf width experience longer leaf cooling time and thus increased thermal time constant compared to leaves with smaller leaf width. Dissections in leaves such as lobes decreases leaf width and thus thermal time constant [@leighInfluenceLeafSize2017]. Additionally, leaf with a smaller time constant experiences variable leaf temperatures in relation to the environment and exhibits instantaneous carbon assimilation rate and stable photosynthesis enabled by greater LMA [@michaletzEnergeticCarbonEconomic2016] Guessing, sun exposed leaves have lower thermal time constant than shaded?      
[I have Michaletz's code to calculate the thermal time constant, so if there are data, or general tendencies for changes in leaf area, LMA, leaf width and leaf dry matter content we can determine how these trends would translate into changes in thermal time constant. 
In practice, an important driver seems to be leaf water content per unit area (I think Lawren might have mentioned this as well in our meeting), so if we can conclude that that is something that scales with the vertical gradient, then we can make an informed guess about how the thermal time constant will change--Martijn Slot]

@@

Leaves with open stomata and high radiation loads maintain $T_{leaf}$ similar to $T_{air}$, with coolest leaves at intermediate sizes (~10mm). Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance. However, counteracting this, lower wind speed in understory would reduce latent heat loss. Thus, under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance. However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday leading to a decrease in leaf CO2 and consequentially increasing leaf temperature in upper canopy [@fauset_differences_2018]. Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.

@@

Supporting the significant negetive relationship between $T50$ and thermal safety margins, photosynthesis [PS2] temperature regulation has been associated leaf thermal tolerance in order to maximize net carbon assimilation. However, randomly collected within-canopy leaves of individual mature plants in Florida showed that Tmax is not directly coordinated with high PS2 heat tolerance and does not necessarily increase leaf thermal tolerance [shows no direct correlation]; nor increases carbon assimilation at higher temperature. Rather, high PS2 heat tolerance reduces optimum temperature for C assimilation [$Topt$] because high PS2 heat tolerance requires a greater cost of PS2 metabolic activity maintainance thus not contributing to increased C assimilation. Species exhibiting low carbon assimilation rate tended to have greater stress tolerance. In addition, species with low stomtal conductance exhibit greater PS2 heat tolerance because earlier closure of stomata ceases C assimaltion before thermal limits are reached [Perez et al. 2020]

>>>>

Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. Overstory trees form canopies that play a crucial role in moderating macro-microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory. This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and Co2 which influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].

The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019], each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019]. In wetter conditions, available water facilitates leaf transpiration through open stomata; however increased relative humdity in tropical regions indirectly caused by increased temperature reduces transpiration, leading to high leaf temperatures in the canopy [Tibbitts 1979; Perez and Feeley 2018].

Frequent drought-related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020]. Larger trees suffer hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al. 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a]. Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020]. However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates, unless relative humidity is so high as to reduce rate of evapotranspiration [Perez and Feeley, 2018].

@@

While there is evidence for temperature driven shifts in forest dynamics, leaf-level metabolic and performance changes influenced by biophysical conditions and warming in the overstory and understory is still unclear. Thus, it is becoming important to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy. This will provide a pathway for understanding how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions with additional feedbacks to forest structure and microclimates.

@@

Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [Rosendaal et al. 2006]. After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [Keenan and Niinemets et al.2016; @niinemetsLeafAgeDependent2016a]. A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR, and the ratio of red to far red radiation [R:FR] intercepting the leaf surface. This enables acclimation or avoidance responses through the development of necessary leaf traits [Casal, 2013]. Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], ultimately influencing leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al. 2016].

@@

Additionally, shade leaves in the understory can be broken into three categories; 1. shade leaves of canopy trees in which the--top of the tree is sun exposed, bottom is shaded. 2. seedlings/saplings of canopy trees that will eventually, if they survive, reach the canopy. 3 understory plants that complete their lifecycle in the shaded understory. These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered [refs, -Martijn Slot]

@@

Taller trees in dense forests tend to be exposed to greater irradiance but maintain shaded leaves in canopy interior and lower canopy, while shorter understory trees are shaded by overstory. As a result of different light regimes, a sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, and hairier, with thicker cuticle and greater lobation than shade leaves
[Rozendaal et al. 2006, Mathur et al. 2018, more REFS]."Additionally, shorter understory trees that spend the bulk of their lifecycle in the shade experience different pressures than plant that get the bulk of their carbon from sun-exposed leaves, but can maintaing a few layers of shade leaves as well" -Martijn

The combination of stressors experienced by taller trees' foliage are different from that experienced by understory trees. For example, overstory sun leaves experience greater loads of PAR, $VPD$, elevated temperatures, higher wind speed, etc.. whereas understory leaves are subjected to shade or limitation of light, niche competition and herbivory. Overstory sun leaves are subjected to overheating pressures; thus, through developed traits of increased stomatal density, transpiration and water supply, along with increased convective heatloss via reduced boundary layer resistance, sun leaves are able to maintain optimum temperatures for photosynthesis and metabolic processes. Additionally, overstory sun leaves tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Adaptive traits in understory enable light capture efficiency while maintaining larger width leaves with comparatively lower stomatal density due to available water, lesser $VPD$, and wind speed. Those differences in traits are likely to affect their thermal properties as well. Leaves act as windows to plant performance, enhancing plant adaptation to changing environment, thus leaf traits acclimated to prior stressors have been suggested to enable whole plants to respond differently to forthcoming stress [Rozendaal et al. 2006; @niinemetsResponsesForestTrees2010a]

@@

With increasing height and age, competition for light decreases as availability  of light increases in dense forests. However, depending on growth strategies of species occupying the understory, leaf traits might differ due to species inherent light requirements. Early successional species are considered more shade intolerant than late successional species that are more shade tolerant. Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$), greater photosynthetic capacity in low light and greater carbon gain capacities in shade tolerators than in intolerators [Niinemets, 2006]. Shade intolerant species experience greater heterogeniety in light during ontogeny [why?-nidhi],thus LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light [H. Poorter, U. Niinemets, L. Poorter, I. J. Wright, R. Villar, Causes and consequences of variation in leaf mass per area (LMA): A meta-analysis. New Phytol. 182, 565–588 (2009).].Rozendaal et al. 2006 found no significant trait differences with ontogeny and adult stature trees. Koike et al. 2001 has more info about this.

@@

Leaf temperature is generally assumed to increase with leaf area ($LA$) [ Wright et al. 2019]. Yet, thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017]. With increase in efffective leaf width, boundary layer thickness and resistance increases along with stomtal resistance [@leighInfluenceLeafSize2017; Wright et al. 2019; @fauset_differences_2018].This lowers evaporation from stomatal pores, and in turn transpirational cooling in large leaves [@fauset_differences_2018]. Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017]. Using leaf energy balance to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al. 2016 found leaf width to be a dominant factor in determining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$. Leaf width and characteristic dimension is generally greater in shaded leaves along the vertical gradient than exposed upper canopy [ref]. Additionally, leaf characteristics such as lobes, leaf elongation and serration, which break the entirety of leaf dimension by enhancing a smaller characteristic dimension, all facilitate more efficient heat dissipation. Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [Vogel 1970; Roth-Nebelsick 2001;Schuepp 1993; Sack et al. 2006].However, Perez and Feeley et al. 2020 showed only a slight negative effect on leaf temperature with leaf width.

@@

Thermal time constant, a leaf trait, quantifies heat flux in a leaf through incorporating heat stored versus heat exchanged with the environment, with relation to time. It comprises of several underlying traits -- leaf area, LMA, leaf width, leaf dry matter content and stomatal conductance [@michaletzEnergeticCarbonEconomic2016]. Typically, larger leaves with greater leaf width experience longer leaf cooling time and thus increased thermal time constant compared to leaves with smaller leaf width. Dissections in leaves, such as lobes, decrease leaf width and thus thermal time constant [@leighInfluenceLeafSize2017]. Additionally, a leaf with a smaller time constant experiences variable leaf temperatures in relation to the environment and exhibits instantaneous carbon assimilation rate and stable photosynthesis enabled by greater LMA [@michaletzEnergeticCarbonEconomic2016] Guessing, sun exposed leaves have lower thermal time constant than shaded?      
[I have Michaletz's code to calculate the thermal time constant, so if there are data, or general tendencies for changes in leaf area, LMA, leaf width and leaf dry matter content we can determine how these trends would translate into changes in thermal time constant. 
In practice, an important driver seems to be leaf water content per unit area (I think Lawren might have mentioned this as well in our meeting), so if we can conclude that that is something that scales with the vertical gradient, then we can make an informed guess about how the thermal time constant will change--Martijn Slot]

@@

Leaves with open stomata and high radiation loads maintain $T_{leaf}$ similar to $T_{air}$, with coolest leaves at intermediate sizes (~10mm). Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance. However, counteracting this, lower wind speed in understory would reduce latent heat loss. Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance. However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday lead to a decrease in leaf CO2, and consequentially increasing leaf temperature in upper canopy [@fauset_differences_2018]. Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.

@@

Supporting the significant negative relationship between $T50$ and thermal safety margins, photosynthesis [PS2] temperature regulation has been associated with leaf thermal tolerance in order to maximize net carbon assimilation. However, randomly collected within-canopy leaves of individual mature plants in Florida showed that Tmax is not directly coordinated with high PS2 heat tolerance and does not necessarily increase leaf thermal tolerance [shows no direct correlation]; nor does it increase carbon assimilation at higher temperature. Rather, high PS2 heat tolerance reduces optimum temperature for C assimilation [$Topt$] because high PS2 heat tolerance requires a greater cost of PS2 metabolic activity maintenance, thus not contributing to increased C assimilation. Species exhibiting low carbon assimilation rates tended to have greater stress tolerance. In addition, species with low stomatal conductance exhibit greater PS2 heat tolerance because earlier closure of stomata ceases C assimilation before thermal limits are reached [Perez et al. 2020]
